TBD

### 2020-11-30 Articles/2020-ClassicalMechanicsPhil/Source/classical-phil-article.tex
Update classical-phil-article.tex
Edits to derivation of action principle and categorical equivalence section

\usepackage{amsmath, hyperref, enumerate, comment}
\usepackage{amsfonts}
\usepackage{graphicx}

\usepackage[backend=biber,  style=chicago-authordate, url=false, doi=false, eprint=false]{biblatex}
%\usepackage[backend=biber,  style=authoryear-icomp, url=false, doi=false, eprint=false]{biblatex}
\addbibresource{bibliography.bib}

@@

Thesis: despite the mathematical equivalence of Hamiltonian and Lagrangian mechanics within the hyper-regular domain, the two frameworks have important differences. In particular, Hamiltonian systems arise when a smaller number of basic physical conditions are met. A Lagrangian system must satisfy a further physical condition. This shows that Hamiltonian mechanics is in a sense more fundamental. Defending this claim does not require commitments to fundamental structure or perfectly natural properties.

@@

Finally, a system satisfies a further condition---motion equivalence---when its trajectory in physical space determines and is determined by its evolution in phase space. This condition leads to a correspondence between velocity and momentum, which formally leads to the framework of Lagrangian mechanics in the hyper-regular domain (Section~\ref{motion}). Hence, Lagrangian mechanics requires that a further physical condition is satisfied, and hence it is less general than Hamiltonian mechanics. To see this, note that some systems fail to satisfy the condition of motion equivalence, i.e. it is not always true that we can determine the evolution of a system in phase space from its trajectory in physical space. For instance, knowledge of the trajectory of a photon is insufficient for knowledge of its phase space evolution, since we cannot determine the momentum of a photon from its trajectory alone. Hence, the set of classical systems that satisfy this further physical condition is a proper subset of the classical systems that satisfy the conditions for Hamiltonian mechanics. On these grounds, we will argue in Section~\ref{fundamentality} that Hamiltonian mechanics is more fundamental than Lagrangian mechanics.

@@

@@

\begin{quotation}
\noindent
\textit{Motion equivalence assumption}: the motion of the system (i.e. trajectories in physical space-time) suffices to recover its dynamics (i.e. evolutions in state space) and vice-versa.
\end{quotation}

@@

@@

@@

Barrett shows that in the hyper-regular domain, Lagrangian and Hamiltonian mechanics are categorically equivalent provided that we define categories of models over the tangent and cotangent bundle formulations, respectively [1181-82]. However, he also shows that if we consider the more general formulation of Hamiltonian mechanics on a symplectic manifold (which does not require a cotangent bundle), then the two formulations are not categorically equivalent [1182-83]. This parallels the construction of Hamiltonian mechanics in Section~\ref{derivation}.

\item Barrett footnote 16 conjectures that outside the hyper-regular domain, Hamiltonian and Lagrangian formulations ``will be inequivalent according to any reasonable standard of equivalence. There is no natural way to translate a non-hyper-regular model of the one theory into a model of the other" [1179]. Does GC framework provide a definite answer on this question of inequivalence?

@@

@@

Possible objection to consider stemming from categorical equivalence: for systems in the hyper regular domain, does the categorical equivalence of Hamiltonian and Lagrangian mechanics show that we require the same number of assumptions to treat systems in this domain?

The answer is presumably ``no", given Barrett's proof that the symplectic formulation of Hamiltonian mechanics is not categorically equivalent to the tangent bundle formulation of Lagrangian mechanics.

Some relevant considerations for responding: note that the photon does not belong to the hyper regular domain. Furthermore, if you are not in the hyper regular domain, then there is no longer a one-to-one map between position and velocity. In this sense, the Hamiltonian framework is more general because it applies outside the hyper regular domain: the Hamiltonian can define a unique solution (unique evolution in phase space) even when the Hessian is zero. In contrast, whenever the Lagrangian is not hyper regular, the Euler--Lagrange equations do not possess a unique solution. Furthermore, density plays a special role in the Hamiltonian framework, whereas it is in some sense an accident that the density in configuration space is invariant (simply because the Jacobian equals unity in Cartesian coordinates).

>>>>

\usepackage{amsmath, hyperref, enumerate, comment}
\usepackage{amsfonts, esint}
\usepackage{graphicx}
\usepackage{mathrsfs} %for script font

\usepackage[backend=biber, style=authoryear, url=false, doi=false, eprint=false]{biblatex}
%\usepackage[backend=biber,  style=authoryear-icomp, url=false, doi=false, eprint=false]{biblatex}
\addbibresource{bibliography.bib}

@@

Thesis: despite the mathematical equivalence of Hamiltonian and Lagrangian mechanics within the hyper-regular domain, the two frameworks have important differences. In particular, Hamiltonian systems arise when a smaller number of basic physical conditions are met. A Lagrangian system must satisfy a further physical condition. This shows that Hamiltonian mechanics is in a sense more fundamental. Defending this claim does not require commitments to fundamental structure or perfectly natural properties. (Furthermore, the Hamiltonian framework is more general because a Hamiltonian defines a unique solution (unique evolution in phase space) even when the Hessian is zero. In contrast, whenever the Lagrangian is not hyper-regular, the Euler--Lagrange equations do not possess a unique solution. Additionally, density plays a special role in the Hamiltonian framework, whereas it is in some sense an accident that the density in configuration space is invariant (simply because the Jacobian equals unity in Cartesian coordinates).

%Note that the photon does not belong to the hyper-regular domain. Furthermore, if you are not in the hyper-regular domain, then there is no longer a one-to-one map between position and velocity.

@@

Finally, a system satisfies a further condition---motion equivalence---when its trajectory in physical space determines and is determined by its evolution in phase space. This condition leads to a correspondence between velocity and momentum, which formally leads to the framework of Lagrangian mechanics in the hyper-regular domain (Section~\ref{motion}). Hence, Lagrangian mechanics requires that a further physical condition is satisfied, and hence it is less general than Hamiltonian mechanics. To see this, note that some systems fail to satisfy the condition of motion equivalence, i.e. it is not always true that we can determine the evolution of a system in phase space from its trajectory in physical space. For instance, knowledge of the trajectory of a photon is insufficient for knowledge of its phase space evolution, since we cannot determine the momentum of a photon from its trajectory alone. In general, different Lagrangians---differing even by more than a constant---can correspond to the same particle trajectories while not agreeing on the total energy, even up to a constant.\footnote{\textcites[1185--1186]{Barrett2} discusses this in the context of the failure of equivalence between the vector field formulations of Lagrangian and Hamiltonian mechanics. In contrast, the trajectories in phase space (encoded by the Hamiltonian vector field) uniquely define the Hamiltonian up to a constant.} Hence, the set of classical systems that satisfy this further physical condition is a proper subset of the classical systems that satisfy the conditions for Hamiltonian mechanics. On these grounds, we will argue in Section~\ref{fundamentality} that Hamiltonian mechanics is more fundamental than Lagrangian mechanics.
[To do: verify that the scenario of two different Lagrangians giving rise to the same trajectories in physical space with different energies, arises only when motion equivalence condition fails. See Curiel 2014, page 42 for his take on this kind of scenario.]

% One way to motivate the motion equivalence condition: in general, the same trajectory in physical space can correspond to different trajectories in phase space: knowledge of the position and velocity is not sufficient for knowledge of the momentum (e.g. if you do not know the mass of the system). Formally, different Lagrangians can have the same Lagrangian vector field but give rise to Hamiltonians with different Hamiltonian vector fields, as Barrett notes \parencites*[1186]{Barrett2}. In contrast, if two Hamiltonian systems agree on the Hamiltonian vector field, then they must agree on the Hamiltonian, up to an arbitrary constant (whereas two Lagrangian systems can have the same Lagrangian vector field but not agree on the Lagrangian or total energy, even up to a constant)).

@@

% One way to motivate the motion equivalence condition: in general, the same trajectory in physical space can correspond to different trajectories in phase space: knowledge of the position and velocity is not sufficient for knowledge of the momentum (e.g. if you do not know the mass of the system). Formally, different Lagrangians can have the same Lagrangian vector field but give rise to Hamiltonians with different Hamiltonian vector fields, as Barrett notes \parencites*[1186]{Barrett2}. In contrast, if two Hamiltonian systems agree on the Hamiltonian vector field, then they must agree on the Hamiltonian, up to an arbitrary constant (whereas two Lagrangian systems can have the same Lagrangian vector field but not agree on the Lagrangian or total energy, even up to a constant)).

@@

% [to do: make the footnote regarding connection to Marsden and Abraham more precise. Might want to flag that their theorem is expressed in terms of the Lagrangian vector field and corresponding Hamiltonian vector field (obtained via Legendre transformation) having the same base integral curves. These base integral curves encode how particles evolve over time. Should I also elaborate on how one can interpret motion equivalence in terms of the theories having the same empirical content? This is how Barrett interprets the Marsden and Abraham theorem 3.6.2]
\begin{quotation}
\noindent
\textit{Motion equivalence assumption}: the motion of the system (i.e. trajectories in physical space-time) suffices to recover its dynamics (i.e. evolutions in state space) and vice-versa.\footnote{ \textcites[]{Abraham1978}, Theorem 3.6.2 shows that in the hyper-regular domain, Lagrangian and Hamiltonian mechanics describe the same evolution of particle positions over time (see \textcites[1180-1181]{Barrett2} for discussion). Motion equivalence can be understood as the converse of this theorem: assuming that Lagrangian and Hamiltonian mechanics agree on particle evolutions, we derive that the Legendre transformation is well-defined in the hyper-regular domain.}
\end{quotation}

@@

\subsection{A geometric interpretation of the action principle}
\label{action}

% Somewhere below, emphasize that you have a unique trajectory only if the action is a minimum or maximum (if it is just stationary but not a minimum or maximum, then you do not have a unique trajectory)

Hamilton's principle of stationary action serves as a foundational principle for Lagrangian mechanics. According to this principle, infinitesimal variations of the action $ \mathscr{A}$ along the system's trajectory $\gamma $ equal zero: $\delta  \mathscr{A}[\gamma] = 0 $. Although central to Lagrangian mechanics, we can actually derive the stationary action principle using Hamiltonian mechanics. Our derivation simply assumes that 1) the system trajectory follows a deterministic and reversible evolution and 2) the system satisfies the motion equivalence condition. We present the derivation for the one-dimensional case---where it is most easily visualized---but it can be generalized using differential geometry.\footnote{For a similar argument illustrating the generalization, see \textcites[110]{Arnold1989}. [To do: verify that Arnold can legitimately be read as generalizing the argument below.]}
% Note on the generalization: cannot use curl. Need to use a form that measures the number of states on a surface. The displacement field kills this form (eats the degenerate direction that does not form degrees of freedom).
% Something that I thought was a second assumption but turns out that this follows just from double role played by time variable: 2) the system's motion always advances in time (so that $dt/dt = 1$).

Knowing that the system is deterministic and reversible, it follows that the density of states is invariant. This means that possible trajectories of the system are never pushed toward each other or pulled apart. In other words, the flux of trajectories through any closed surface is always zero. Formally, the displacement of the states in time is divergenceless: $\nabla \cdot \vector{S} = 0 $. Here, $\vec{S} := (\frac{d q }{d t }, \frac{d p }{d t }, \frac{d t }{d t })$ defines a vector field in the extended phase space $(q, p , t ) $. $\vector{S}$ characterizes the displacement of states in time. In particular, it is tangent to the trajectory $\gamma [q, p, t] (t)$, in extended phase space, where the variable $t$ functions as both a coordinate and affine parameter of the trajectory.

Since $\vector{S}$ is divergenceless, there exists another vector field $-\vector{\theta}$ such that $\vector{S} = curl(-\vector{\theta} )$.\footnote{For reasons that will become clear, we choose $-\vector{\theta}$ rather than $\vector{\theta}$.} As a vector field in the extended phase space, $\vector{\theta}$ can be written in terms of its components $\vector{\theta} = (\theta_q, \theta_p, \theta_t) $. Since the curl of the gradient of a scalar function is always zero, the displacement vector $\vector{S}$ is invariant under the gauge transformation $\vector{\theta'} =\vector{\theta} + grad(f)$, where $f$ is an arbitrary scalar function. Hence, we can choose $f$ such that $\partial f/ \partial p = -\theta_p $. Without loss of generality, we can thereby redefine $\vector{\theta}$ as $(\theta_q, 0, \theta_t) $.

Treating time $t$ as both a coordinate and an affine parameter (parameterizing the trajectory), the time component of the displacement, $S_t = dt/ dt = 1$.\footnote{Considering $dt/dt$, the time variable in the numerator functions as a coordinate, while the time variable in the denominator functions as a parameter. We could instead use a parameter $\tau$, and the derivation would remain valid.} This places another constraint on $\vector{\theta} $, since $\vector{S} = -curl(\vector{\theta}) $. In components, this means that $S_t = - (\frac{\partial}{\partial q} \cdot \theta_p - \frac{\partial}{\partial p} \cdot \theta_q) = - (\frac{\partial}{\partial q} \cdot 0 - \frac{\partial}{\partial p} \cdot \theta_q) = \frac{\partial q}{\partial p}$. Hence, $S_t = 1$ entails that $\frac{\partial q}{\partial p} = 1$. Integrating this expression, we find that $\theta_q = p$ up to addition by an arbitrary constant function $g(q)$. Again using the gauge freedom of $\vector{S}$ we can set $g(q)=0$, so that $\theta_q = p$. Thus, so far, we have shown that $\vector{\theta} = (p, 0,\theta_t) $.

Next, we introduce a function $H$ and let $ \theta_t \equiv -H$. Hence, we write $-\vector{\theta} = (-p, 0, H) $ as the vector potential for the displacement vector $\vector{S} $. By matching components of $\vector{S}$ and $-curl(\theta)$, we show that H is the Hamiltonian for the system. First, by definition, $S_q \equiv \frac{dq}{dt} $ and $S_p \equiv  \frac{dp}{dt} $. Secondly, since $\vector{S} = -curl(\vector{\theta}) $, we calculate that $S_q = -\frac{\partial \theta_t}{\partial p} = \frac{\partial H}{\partial p}$ and $S_p = \frac{\partial \theta_t}{\partial q}=-\frac{\partial H}{\partial q}$. Matching terms, we see that $\frac{\partial H}{\partial p} =\frac{dq}{dt} $ and $-\frac{\partial H}{\partial q} =\frac{dp}{dt}$. These simply are Hamilton's equations, which justifies interpreting $H$ as the Hamiltonian.

% Since the system is deterministic and reversible, it satisfies Hamilton's equations: $\frac{\partial H}{\partial p} =\frac{dq}{dt} \equiv S_q$ and $-\frac{\partial H}{\partial q} =\frac{dp}{dt} \equiv S_p $. Additionally, since $\vector{S} = -curl(\vector{\theta}) $, we know that $S_q = -\frac{\partial \theta_t}{\partial p}$ and $S_p = \frac{\partial \theta_t}{\partial q}$. Integrating the first equation for $\theta_t$ gives $-H$ up to an arbitrary constant function of q, while the second gives $\theta_t = -H$ up to an arbitrary constant function of p. For consistency, these two constant functions must vanish, so that $\theta_t = -H$ [check argument here; do we actually need to appeal to the gauge freedom of S again?]. Hence, it follows that we can take $-\vector{\theta} = (-p, 0, H) $ as the vector potential for the displacement vector $\vector{S} $.

At this point, we return to the action $ \mathscr{A}$, defined as the integral of the Lagrangian over a trajectory beginning at time $t_0$ and ending at time $t_1$: $\integral^{t_1}_{t_0} L (q, \dot{q}, t) dt$. Provided that the Legendre transform is well-defined (i.e. the motion equivalence assumption is satisfied), then $H = p \dot{q} - L$. Hence, we can re-express the action as $\integral^{t_1}_{t_0} L (q, \dot{q}, t) dt = \integral^{t_1}_{t_0} [p \frac{dq}{dt} - H(q, p, t) ]dt$. Multiplying the Hamiltonian by $dt/dt = 1$ and adding a vanishing $0 \cdot \frac{dp}{dt}$, we obtain $ \mathscr{A} = \integral^{t_1}_{t_0} [p \frac{dq}{dt} + 0 \cdot \frac{dp}{dt} - H(q, p, t) \frac{dt}{dt} ]dt$. We notice that the integrand simply equals the inner product of $\vector{\theta}$ with $\vector{S} $, entailing that the Lagrangian $L =\vector{\theta} \cdot \vector{S} $. Consequently, the action $ \mathscr{A} [\gamma] = \integral^{t_1}_{t_0} \vector{\theta} \cdot \vector{S}  dt$. Changing variables, we take the line integral of $\vector{\theta}$ along the trajectory $\gamma $: $ \mathscr{A} [\gamma] =\integral_{\gamma} \vector{\theta} \cdot d\vector{\gamma}$.\footnote{Note that $\vector{S}  dt = d\vector{\gamma}$ only on the actual trajectory in phase space, but not on other possible trajectories.} To derive Hamilton's principle of stationary action, it suffices to show that $\delta \mathscr{A} [\gamma] =\delta \integral_{\gamma} \vector{\theta} \cdot d\vector{\gamma} = 0$. 
% Note on the generality of the argument: in cases where motion equivalence is not satisfied, we can still write the integrand in terms of H(q, p, t), although it would not be an integral over the Lagrangian. Thus, we can still express the `action' as a function of p, q, t. We would calculate this on a path in the extended phase space, although without motion equivalence, not all of the paths on phase space would correspond to a unique trajectory in real, physical space. See if Arnold has a more general name for this generalized principle of stationary action.

By definition, a variation over the integral $\integral_{\gamma} \vector{\theta} \cdot d\vector{\gamma}$ consists in taking the difference between the integrand and a corresponding integral over a nearby trajectory $\gamma'$: $\delta \integral_{\gamma} \vector{\theta} \cdot d\vector{\gamma} = \integral_{\gamma} \vector{\theta} \cdot d\vector{\gamma} - \integral_{\gamma'} \vector{\theta} \cdot d\vector{\gamma'}$ . Since the trajectories $\gamma$ and $\gamma'$ have the same endpoints, they define a closed-curve $C$, enclosing a region $\Sigma $. Hence, by Stoke's theorem $\delta \integral_{\gamma} \vector{\theta} \cdot d\vector{\gamma} = \oint_{C} \vector{\theta}  \cdot dC = \iint_{\Sigma} curl(\vector{\theta}) \cdot \diff \vector{\Sigma}$ . Recalling now that 1) the $curl(\vector{\theta})$ equals the negative of the displacement vector $\vector{S}$, and 2)  $\vector{S}$ is tangent to the trajectory $\gamma $, we see that $\vector{\theta}$ is tangent to the surface $\Sigma $.\footnote{Since $\vector{S}$ is tangent to $\gamma $, and $\gamma'$ defines an infinitesimal variation, it follows that $\vector{S}$ is tangent to the entire surface $\Sigma $.} Hence, the surface integral must vanish, showing that $\delta \integral_{\gamma} \vector{\theta} \cdot d\vector{\gamma}$ does indeed equal zero. This establishes Hamilton's principle of stationary action, which we have derived by assuming that 1) the displacement vector is divergenceless and 2) using the Legendre transform (assuming motion equivalence).

@@

% Helpful discussion of differences between Riemannian and symplectic manifolds from Wikipedia (https://en.wikipedia.org/wiki/Symplectomorphism): ``Unlike Riemannian manifolds, symplectic manifolds are not very rigid: Darboux's theorem shows that all symplectic manifolds of the same dimension are locally isomorphic. In contrast, isometries in Riemannian geometry must preserve the Riemann curvature tensor, which is thus a local invariant of the Riemannian manifold. Moreover, every function H on a symplectic manifold defines a Hamiltonian vector field XH, which exponentiates to a one-parameter group of Hamiltonian diffeomorphisms. It follows that the group of symplectomorphisms is always very large, and in particular, infinite-dimensional. On the other hand, the group of isometries of a Riemannian manifold is always a (finite-dimensional) Lie group. Moreover, Riemannian manifolds with large symmetry groups are very special, and a generic Riemannian manifold has no nontrivial symmetries."

@@

Barrett shows that in the hyper-regular domain, Lagrangian and Hamiltonian mechanics are categorically equivalent provided that we define categories of models over the tangent and cotangent bundle formulations, respectively \parencites*[1181-82]{Barrett2}. However, he also shows that if we consider the more general formulation of Hamiltonian mechanics on a symplectic manifold (which does not require a cotangent bundle), then the two formulations are not categorically equivalent \parencites*[1182-83]{Barrett2}. This parallels the construction of Hamiltonian mechanics in Section~\ref{derivation}.

% Not sure best place to put the following, if at all:
Nevertheless, based on the equivalence between the tangent bundle and cotangent bundle formulations of Lagrangian and Hamiltonian mechanics, one might wonder whether there could be a ``correspondingly general formulation of Lagrangian mechanics" that is analogous to the symplectic but-not-cotangent-bundle formulation of Hamiltonian mechanics \parencites[1182]{Barrett2}. For, if there were, then this formulation of Lagrangian mechanics might be more general than the cotangent bundle formulation of Hamiltonian mechanics. And thus, there would be an analogous sense in which Lagrangian mechanics is more fundamental than Hamiltonian mechanics, undercutting our argument. Fortunately, it is implausible that there could be a suitable generalization of Lagrangian mechanics. In going from the cotangent bundle formulation of Hamiltonian mechanics to the symplectic manifold formulation, we lose some globally-defined structure. This is analogous to moving from the space of globally-Euclidean manifolds to the space of locally-Euclidean manifolds. Yet, it is not clear how Lagrangian mechanics could remain well-defined while throwing away global structure from the tangent bundle. For the action seems to be inherently a globally-defined quantity, as an integral defined over a trajectory.

[To do: not sure if following warrants discussion:]
Teh and Tsementzis \parencites*[]{Teh} provide another argument for the equivalence of Lagrangian and Hamiltonian mechanics within the hyper-regular domain. They develop this using Tulczyjew's reformulation of Hamiltonian mechanics, which shows that ``the very concept of a classical mechanical system (and not merely the phase space of that system) should be described in terms of symplectic geometry" \parencites*[46]{Teh}. Again, there is a worry that this equivalence result might undercut our privileging of Hamiltonian mechanics. However, the equivalence is again just within the hyper-regular domain, requiring that motion equivalence is satisfied. Furthermore, Tulczyjew's formulation relies heavily on the notion of a Lagrangian submanifold, since ``a mechanical system turns out to be a Lagrangian submanifold of an appropriately defined space" \parencites[47]{Teh}.\footnote{A submanifold $N$ of a symplectic manifold $(P, \omega) $ is Lagrangian provided that i) it has half the dimension of P and ii) the pullback of the symplectic form $\omega$ vanishes along the inclusion of N in P \parencites[48]{Teh}.} Yet importantly, Lagrangian submanifolds do not distinguish between the space of velocities and the space of momenta, i.e. they obscure the important physical differences between these two kinds of quantities. Not every constructible Lagrangian submanifold defines units in a physically significant way.\footnote{One could even construct a suitable Lagrangian submanifold by mixing position and momenta coordinates.} It is the physical significance of defining a unit system that breaks the formal (syntactic) symmetry between position and momentum coordinates. 
% [To do: verify the veracity of this discussion of Lagrangian sub manifolds]
%Furthermore, within this reformulation, Lagrangian submanifolds play a very important role. Does this lead to a possible privileging of Lagrangian mechanics?

[To do: Not sure where to put the following material, if anywhere; perhaps after a discussion of why we restrict to hyper-regular domain?:]
In a footnote, Barrett conjectures that outside the hyper-regular domain, Hamiltonian and Lagrangian formulations ``will be inequivalent according to any reasonable standard of equivalence. There is no natural way to translate a non-hyper-regular model of the one theory into a model of the other" \parencites*[1179]{Barrett2}. Indeed, outside the hyper-regular domain, the Euler--Lagrange equations do not have a single, unique solution. Hence, it is difficult to say what a Lagrangian system even is outside the hyper-regular domain. As an example, consider the case where the Lagrangian $L$ equals zero. Then for this ``system," all paths have the same action, so all paths are equally possible.\footnote{This Lagrangian corresponds to a Hamiltonian that also equals zero, entailing that $dH = 0 $. By Hamilton's equations, the derivatives of the position and momenta are zero, indicating that this system is static.} In contrast,  a Hamiltonian defines a unique solution (unique evolution in phase space) even when outside the hyper-regular domain (i.e. when the Hessian is zero).

\begin{comment}

Notes/questions from 11-16-20 meeting
\begin{enumerate}[1.)]

@@

\item: GC in response: defining the Lagrangian presupposes that we have trajectories of positions over time. Velocity is then a derived quantity of these positions. So the tangent bundle seems to be intimately connected with having trajectories. Hence, if we eliminated reference to the tangent bundle, we would seemingly not be able to define trajectories.

\item Furthermore, there is a sense in which GC derivation recovers the tangent bundle and shows why it must be well-defined: provided the evolutions are differentiable (which is needed to preserve densities), it follows that velocities are well-defined, and hence so is the tangent bundle. Note that the trajectories alone do not entail that velocities are well-defined, since the trajectories could in principle not be differentiable.

\item Note that although the symplectic manifold already has some global structure (which corresponds to the global tangent bundle), the cotangent bundle itself is in some sense more global (and hence more specialized). Helpful analogy: a manifold can be locally Euclidean (which is a global future) without being globally Euclidean. The sense in which the tangent bundle and symplectic manifold structure are global is analogous to being locally Euclidean, whereas the global cotangent bundle structure is analogous to being globally Euclidean. So in going from the cotangent formulation to the symplectic formulation, we are becoming less specific (more general). To generalize the Lagrangian, we would seemingly need to lose some structure, but the action seems to be global (integral defined over a trajectory).

\item It might be worth noting these differences, as a way of motivating the claim that there probably is not a correspondingly general formulation of Lagrangian mechanics.

@@
